Geometric variational problems

### 2020-11-12 NSF2020/Description/specificaim1_ver2.tex
yny specific aim 1

Hamm and Kozlov (HK) pioneered 
the modern theory of membrane continuum mechanics \cite{Hamm2000}, and their theory
%was pioneered by Hamm and Kozlov (HK) \cite{Hamm2000}. The theory 
is widely used to describe biological phenomona 
fission \cite{FrEsAkSh15, Maetal15, PhysRevE.79.031926},
fusion \cite{ChKo08, KoKo2002,Kuzmin7235,Aeffner2012},
poration \cite{Gaetal20}, phase boundaries and interaction with inclusions
\cite{SeLeMaEg17,Saetal20, Pietal20}, which require resolution of the internal structure of the membrane.  
%Using the HK  model as a base,
%PI RR and collaborators calculated, for the first time in continuum theory, a least energy path for transitions between planar bilayers, a membrane stalk,
%hemifusion diaphragm and the fusion pore \cite{RyWaCo13,RyKlYaCo16},
%and the energies calculated by these continuum studies (Figure \ref{fig:barriers}A),
%are in agreement with barrier heights derived by molecular dynamics and experimental studies \cite{FrRoPi17}.  
%
Recently, there has been a revival in interest in the HK theory as the quadratic assumption
for the elasticity energy density has caused researchers to question the applicability of \eqref{ansatz3} for large curvatures \cite{PhysRevLett.117.188102, ARGUDO20161619}. 
For example, Galimzyanov {\it et al.} \cite{C9SM02079A} countered that energies derived from molecular dynamics and elasticity agree, even when curvatures are large.
This proposal will develop much-needed mathematical analysis to resolve these controversies due to the assumptions in the HK theory.

\begin{figure}
  \centering{\includegraphics[width=0.80\textwidth]{Figures/Deformations.pdf}}
  \caption{\label{fig:deformations}Sketch of the HK membrane model \cite{Hamm2000}.
    }
\end{figure}

@@

%\begin{center}
%Anticipated model simulations
%\end{center}
\subsubsection{Anticipated model simulations}
To characterize material properties of a bilayer, we compare the energies and forces generated by
a collection of amphiphilic particles against the energies and forces of a continuum bilayer.
We propose to first {\bf (i)} Assess whether the continuum energies \eqref{ansatz3} are the same
as the corresponding energies in the particle system for all reasonable deformation.
If this is the case we will {\bf (ii)} determine the value of the elastic $\KB$,
$\KT$, $\KG$ and $\KTH$, and then {\bf (iii)} understand how the model parameters $\rho,$ $\gamma$, and particle shape map onto elastic moduli. 
%
%Motivated by our preliminary results \cite{Fu2018_SIAM}
%we use realistic values for particle diameter $2$ nm \cite{Boal},
%screening length $\rho = 2.5$ nm \cite{Eriksson1989,Lin2005,Parsegian,Israelachvili80,TerziDeserno17}
%and interfacial tension $\gamma=4.1$ pN nm$^{-1}$ \cite{GarciaSaez, KUZMIN2005, Petelska2012, Jackson}
%model parameters. The repulsion strength, repulsion length scale, particle shape (e.g. disks, ellipses), 
%particle diameter and the hydrophobic boundary condition \eqref{SL}
%are additional parameters defining the coarse-grained representation.

An accurate and robust way to measure material properties is to track the evolution of the bilayer. 
Here we start with a particle-based bilayer in a non-equilibrium shape, and then evolve the particle bilayer by steepest gradient descent with respect to the
HARP functional:
\begin{equation}
\label{SGD}   
\frac{\dif {\bf a}_i}{\dif t} = {\bf F}_i,    \quad  \frac{\dif {\bf R}_i}{\dif t}  = {\bf R}_i A({\bf w}_i),\quad i = 1, \dots, N.
\end{equation}
where that particle centers ${\bf a}_i$ and rotation matrices ${\bf R}_i$ determine the particle configuraton; 
$A({\bf w}_i)$ is the skew-symmetric matrix with axial vector ${\bf w}_i$.
We solve \eqref{SGD} numerically 
using an Adams–Bashforth linear multistep method. The procedure involves solving \eqref{SL} based off of the current particle configuration.
We then use the solution $u$ to calculate the collection of forces and torques \eqref{forceandtorque}, and update the particle configuration
through \eqref{SGD}.

The evolution \eqref{SGD} satisfies the energy dissipation law $\frac{\dif }{\dif t} \Phi \leq 0$
and this stabilizes the evolving bilayer shape.
Therefore, we reconstruct an evolving monolayer dividing surface $\Sigma$ and director field ${\bf n}$ by
interpolating the particle centers and orientations. Using \eqref{ansatz3}, we calculate a continuum energy $E$ from the interpolated shapes.

In certain instances, the bilayer evolution involves only one or a few of the four fundamental deformations.
When this occurs, the elastic moduli derive from the coefficients of the linear regression between the HAP energy $\Delta \Phi = \Phi - \Phi_0$
and continuum energies and $\Delta E = E - E_0$, where  $\Phi_0$ and $E_0$ are respective equilibrium values.

Since it is a particle-based method, the HAP model has a number of advantages over continuum-based descriptions.
The elastic energy \eqref{ansatz3} couples director gradients to surface geometry through
the tilt vector field.  Furthermore, bilayer energy is the sum of two monolayer
energies and the distance between these monolayers varies according to \eqref{LMdeformation}.
Effects like tilt and varying monolayer thickness appear at lipid domain boundaries, the edge of nanopores and at the boundary of membrane inclusions \cite{PhysRevE.102.042406}.
These effects are present in the proposed formulation, and the particle-based bilayers can break open and join monolayer-wise 
as required by fussion and fission.

Bending is the energetically most consequential deformations. To measure a bending modulus,  
we deform the midplane of a planar bilayer into the shape of a circular arc.
This arc stores bending energy in the form of splay $\Div {\bf n}$, while twist, saddle-splay and tilt deformations are zero.
The splay in the two leaflets approximately cancel when added, making the contribution of spontaneous curvature negligible.
In the case of a rectangular bialyer, the corresponding elastic energy 
$E = \int_{\Sigma} \KB \kappa^2$ where $\kappa$ is the curvature of the curved cross-section $C$.
Figure \ref{} shows the approximately linear relationship between $\Delta \Phi$ and $\Delta E$
used to values for infer $\KB$.

The twist deformation is a fully three-dimensional deformation. 
(Specific Aim 3 address outstanding implementation issues like three-dimensional boundary integral solvers.) 
It is closely related to the assumption of lateral fluidity in membranes \cite{Hamm2000}. Note that tilt is locally non-zero whenever twist is non-zero.
That is because the surface gradient $\nabla {\bf n}$ equals the second fundamental form, and is therefore symmetric, whenever ${\bf n} = {\bf N}$ locally.

To measure the twise modulus, we place a single monolayer on a hydrophobic substrate 
and take the initial configuration ${\bf n} = (0, f(x_1), \sqrt{1 - f^2(x_1)})$,
yielding a surface gradient $\nabla {\bf n} = \begin{pmatrix}0 & 0\\ f'(x_1) & 0\end{pmatrix}$. 
  Here $\Div {\bf n} = \mathrm{tr}(\nabla {\bf n}) = 0$, $\det(\nabla {\bf n}) = 0$ and
  $\Curl {\bf n} = (\nabla u)_{12} - (\nabla u)_{21} = f'(x_1)$.
  This gives the continuum elastic energy equals
  \begin{equation}
    \label{eq:twist_energy}
    E = \int_{\Sigma} \tfrac{1}{2}\KT (f'(x_1))^2 + \tfrac{1}{2} \KTH f^2(x_1) \,\dif dx_1 \dif x_2,
  \end{equation}
  which we compare to the HAP simulation data.

We will also determine the effective saddle splay modulus $\KG$. 
The gradient descent technique is ineffective for measuring $\KG$ because
the saddle splay energy is largely invariant under shape changes.
In the simplified Canham-Helfrich formulation, saddle splay energy is an indicator of topological transitions, thanks to the  Gauss-Bonnet theorem
\cite{TerziDeserno17}.
PI-Ryham showed that saddle splay acts as a topological indicator even in the presence of nonzero tilt \cite{RyKlYaCo16}.

\begin{wrapfigure}[16]{r}{0.32\textwidth}
\centerline{\includegraphics[width=0.32\textwidth]{Figures/SaddleSplayDiagram.pdf}}
\caption{
\label{fig:saddle_splay}  
 TBD
}
\end{wrapfigure}
To evaluate $\KG$, we will combine the string method from PI-Ryham's work on fusion \cite{RyKlYaCo16} with
the present particle simulations. The string method is a numerical scheme that finds
least energy pathways separating energy basins \cite{doi:10.1063/1.2720838}.
Using the string method, we will evaluate the transition energies between the intact pancake shape,
and one with a pore. Pore formation in a single bilayer decreases the integral of saddle splay from $4\pi$
to $0$. We can, in principle, detect this drop in energy because we have already accounted for the other deformations.

%We will also analyze $\KG$ using a helical geometry. In a helicoid with inclination $m$, the director
%gradient is $\nabla n = \begin{pmatrix} 0 & m(m^2+x_1^2)^{-1} \\ m(m^2+x_1^2)^{-1} & 0\end{pmatrix}$ where $x_1$ is the distance to the central
%axis. We observe that saddle splay, twist and tilt are zeros, and $\det \nabla n = m^2(m^2+x_1^2)^{-2}$.
%In the simulations, the particles forming the helicoid will be confined to two, concentric cylinders. The
%continuum energy $\tfrac{1}{2}\KG h(m x_1(m^2+x_1^2)^{-1} + \tan^{-1}(x_1/m)) |_{r_1}^{r_2}$ where $h$, $r_1$ and $r_2$
%are the cylinder heights, inner cylinder radius, and outer cylinder radius, respectively.

\subsubsection{New mathematical analysis}

@@

Analogously, the HAP energy converges to the surface energy in the limit of vanishing screening length 
and the boundary layers of the minimizers (the solutions to the screen Laplace equation) converges to 
area measures supported on the particle boundaries. 
PI RR has worked directly on phase field approximations of membrane bending energy 
\cite{0951-7715-18-3-016} and asymptotcs in the Poisson-Boltzman equation \cite{1531-3492_2006_2_357,Lee2018}.
The proposed analysis is a natural outgrowth or prior work because the linearization of the Poisson-Boltzmann equation is the screened Laplace equations \eqref{SL}.
One of the basic questions is to establish the HAP convergence rate as a function of particle geometry 
\cite{LuMo89}, which is an essential benchmark for numerical validation. 
In the context of microstructure fabrication, the particle configurations maximally sequester hydrophobic surfaces.  
As an illustration, consider two fluid-bound plates $P_1$ and $P_2$ that geometrically flat and hydrophobic all sides.
We conjecture that in the zero-screening length limit, the plates have overlapping faces, and that 
\begin{equation}
\lim_{\rho \to 0} \min \Phi_{\text{hydro}} = \max_{\mathcal{X}} A(P_1 \Delta P_2)
\end{equation}
$A(P_1 \Delta P_2)$ is the area of the symmetric difference of the overlap.

>>>>

Hamm and Kozlov (HK) pioneered 
the modern theory of membrane continuum mechanics \cite{Hamm2000}, and their theory
%was pioneered by Hamm and Kozlov (HK) \cite{Hamm2000}. The theory 
is widely used to describe biological phenomona 
fission \cite{FrEsAkSh15, Maetal15, PhysRevE.79.031926},
fusion \cite{ChKo08, KoKo2002,Kuzmin7235,Aeffner2012},
poration \cite{Gaetal20}, phase boundaries and interaction with inclusions
\cite{SeLeMaEg17,Saetal20, Pietal20}, which require resolution of the internal structure of the membrane.  
Recently, there has been a revival in interest in the HK theory as the quadratic assumption
for the elasticity energy density has caused
%Using the HK  model as a base,
%PI RR and collaborators calculated, for the first time in continuum theory, a least energy path for transitions between planar bilayers, a membrane stalk,
%hemifusion diaphragm and the fusion pore \cite{RyWaCo13,RyKlYaCo16},
%and the energies calculated by these continuum studies (Figure \ref{fig:barriers}A),
%are in agreement with barrier heights derived by molecular dynamics and experimental studies \cite{FrRoPi17}.  
%
\begin{wrapfigure}[12]{l}{0.47\textwidth}
\centerline{\includegraphics[width=0.47\textwidth]{Figures/Deformations.pdf}}
%  \centering{\includegraphics[width=0.80\textwidth]{Figures/Deformations.pdf}}
\caption{\label{fig:deformations}Sketch of the HK membrane model \cite{Hamm2000}.
    }
\end{wrapfigure}
%
 researchers to question the applicability of \eqref{ansatz3} for large curvatures \cite{PhysRevLett.117.188102, ARGUDO20161619}. 
For example, Galimzyanov {\it et al.} \cite{C9SM02079A} countered that energies derived from molecular dynamics and elasticity agree, even when curvatures are large.
This proposal will develop much-needed mathematical analysis to resolve these controversies due to the assumptions in the HK theory.

@@

%\begin{center}
%Anticipated model simulations
%\end{center}
\subsubsection{Simulations of the HARP model to estimate elastic moduli}
%To characterize material properties of a bilayer, we compare the energies and forces generated by
%a collection of amphiphilic particles against the energies and forces of a continuum bilayer.
We propose to
%
%to examine conduct calculations similar to the example of figure~\ref{fig:flattening}
%for a wide range of particle parameters and physical parameters in HARP to 
%
%{\bf (i)} Assess whether the continuum energies \eqref{ansatz3} are the same
%as the corresponding energies in the particle system for all reasonable deformation.
%If this is the case we will 
%
{\bf (i)} determine the value of the elastic moduli $\KB$,
$\KT$, $\KG$ and $\KTH$, and then {\bf (ii)} understand how the model parameters $\rho,$ $\gamma$, and particle shape map onto elastic moduli. 
An accurate and robust way to measure material properties is to track the evolution of the bilayer as
%
\begin{wrapfigure}[10]{r}{0.4\textwidth}
\vspace{-0.3cm}
\centerline{\includegraphics[width=0.4\textwidth]{Figures/Flattening.pdf}}
%  \centering{\includegraphics[width=0.80\textwidth]{Figures/Deformations.pdf}}
\vspace{-0.2cm}
\caption{\label{fig:flattening}Example of computing bending modulus of a lipid bilayer from particle simulation.
    }
\end{wrapfigure}
it relaxes from an initially non-equilibrium configuration.
Figure~\ref{fig:flattening} is an example where the initial configuration of a lipid bilayer patch contains only splay deformation without any other components (saddle splay, twist and tilt). 
As the bent membrane flattens we find that both the membrane self-interaction energy $\Phi$ and the elastic energy ${\cal W}$ decrease, and the slope gives the
splay modulus in this case.

%To characterize material properties of a bilayer, we compare the energies and forces generated by
%a collection of amphiphilic particles against the energies and forces of a continuum bilayer.
%We propose to first {\bf (i)} Assess whether the continuum energies \eqref{ansatz3} are the same
%as the corresponding energies in the particle system for all reasonable deformation.
%If this is the case we will {\bf (ii)} determine the value of the elastic $\KB$,
%$\KT$, $\KG$ and $\KTH$, and then {\bf (iii)} understand how the model parameters $\rho,$ $\gamma$, and particle shape map onto elastic moduli. 
%%
%%Motivated by our preliminary results \cite{Fu2018_SIAM}
%%we use realistic values for particle diameter $2$ nm \cite{Boal},
%%screening length $\rho = 2.5$ nm \cite{Eriksson1989,Lin2005,Parsegian,Israelachvili80,TerziDeserno17}
%%and interfacial tension $\gamma=4.1$ pN nm$^{-1}$ \cite{GarciaSaez, KUZMIN2005, Petelska2012, Jackson}
%%model parameters. The repulsion strength, repulsion length scale, particle shape (e.g. disks, ellipses), 
%%particle diameter and the hydrophobic boundary condition \eqref{SL}
%%are additional parameters defining the coarse-grained representation. 
%
%An accurate and robust way to measure material properties is to track the evolution of the bilayer.

Here we start with a particle-based bilayer in a specific non-equilibrium shape that consists of only one component of the displacement in figure~\ref{fig:deformations}, 
and then evolve the particle system
%bilayer by steepest gradient descent 
with respect to the HARP functional:
\begin{equation}
\label{SGD}   
\frac{\dif {\bf a}_i}{\dif t} = {\bf F}_i,    \quad  \frac{\dif {\bf R}_i}{\dif t}  = {\bf R}_i A({\bf w}_i),\quad i = 1, \dots, N.
\end{equation}
where that particle centers ${\bf a}_i$ and rotation matrices ${\bf R}_i$ determine the particle configuraton; 
$A({\bf w}_i)$ is the skew-symmetric matrix with axial vector ${\bf w}_i$.
%We solve \eqref{SGD} numerically  using an Adams–Bashforth linear multistep method. 
The procedure involves solving \eqref{SL} based off of the current particle configuration.
We then use the solution $u$ to calculate the collection of forces and torques \eqref{forceandtorque}, and update the particle configuration
through \eqref{SGD}.

\begin{wrapfigure}[12]{l}{0.32\textwidth}
\centerline{\includegraphics[width=0.32\textwidth]{Figures/SaddleSplayDiagram.pdf}}
\caption{
\label{fig:saddle_splay}  
Determination of saddle splay modulus.
}
\end{wrapfigure}
%
The evolution \eqref{SGD} satisfies the energy dissipation law $\frac{\dif }{\dif t} \Phi \leq 0$
and this stabilizes the evolving bilayer shape.
Therefore, we reconstruct an evolving monolayer dividing surface $\Sigma$ and director field ${\bf n}$ by
interpolating the particle centers and orientations. Using \eqref{ansatz3}, we calculate a continuum energy $E$ from the interpolated shapes.

%In certain instances, the bilayer evolution involves only one or a few of the four fundamental deformations.
%When this occurs, the elastic moduli derive from the coefficients of the linear regression between the HAP energy $\Delta \Phi = \Phi - \Phi_0$
%and continuum energies and $\Delta E = E - E_0$, where  $\Phi_0$ and $E_0$ are respective equilibrium values.

%
%Since it is a particle-based method, the HAP model has a number of advantages over continuum-based descriptions.
%The elastic energy \eqref{ansatz3} couples director gradients to surface geometry through
%the tilt vector field.  Furthermore, bilayer energy is the sum of two monolayer
%energies and the distance between these monolayers varies according to \eqref{LMdeformation}.
%Effects like tilt and varying monolayer thickness appear at lipid domain boundaries, the edge of nanopores and at the boundary of membrane inclusions \cite{PhysRevE.102.042406}.
%These effects are present in the proposed formulation, and the particle-based bilayers can break open and join monolayer-wise 
%as required by fussion and fission. 
%
%Bending is the energetically most consequential deformations. To measure a bending modulus,  
%we deform the midplane of a planar bilayer into the shape of a circular arc.
%This arc stores bending energy in the form of splay $\Div {\bf n}$, while twist, saddle-splay and tilt deformations are zero.
%The splay in the two leaflets approximately cancel when added, making the contribution of spontaneous curvature negligible.
%In the case of a rectangular bialyer, the corresponding elastic energy 
%$E = \int_{\Sigma} \KB \kappa^2$ where $\kappa$ is the curvature of the curved cross-section $C$.
%Figure \ref{} shows the approximately linear relationship between $\Delta \Phi$ and $\Delta E$
%used to values for infer $\KB$.

We propose to conduct calculations similar to the example of figure~\ref{fig:flattening} for other elastic moduli such as the effective saddle splay modulus $\KG$:
%for a wide range of particle parameters and physical parameters in HARP to 
%
%We will also determine the effective saddle splay modulus $\KG$. 
The gradient descent technique is ineffective for measuring $\KG$ because
the saddle splay energy is largely invariant under shape changes.
In the simplified Canham-Helfrich formulation, saddle splay energy is an indicator of topological transitions, thanks to the  Gauss-Bonnet theorem
\cite{TerziDeserno17}.
PI-RR showed that saddle splay acts as a topological indicator even in the presence of nonzero tilt \cite{RyKlYaCo16}.

To evaluate $\KG$, we will combine the string method from PI-Ryham's work on fusion \cite{RyKlYaCo16} with
the present particle simulations. 
%The string method is a numerical scheme that finds
%least energy pathways separating energy basins \cite{doi:10.1063/1.2720838}.
%Using the string method, we will evaluate the transition energies between the intact pancake shape,
%and one with a pore. Pore formation in a single bilayer decreases the integral of saddle splay from $4\pi$
%to $0$. We can, in principle, detect this drop in energy because we have already accounted for the other deformations. 
%
%%We will also analyze $\KG$ using a helical geometry. In a helicoid with inclination $m$, the director
%%gradient is $\nabla n = \begin{pmatrix} 0 & m(m^2+x_1^2)^{-1} \\ m(m^2+x_1^2)^{-1} & 0\end{pmatrix}$ where $x_1$ is the distance to the central
%%axis. We observe that saddle splay, twist and tilt are zeros, and $\det \nabla n = m^2(m^2+x_1^2)^{-2}$.
%%In the simulations, the particles forming the helicoid will be confined to two, concentric cylinders. The
%%continuum energy $\tfrac{1}{2}\KG h(m x_1(m^2+x_1^2)^{-1} + \tan^{-1}(x_1/m)) |_{r_1}^{r_2}$ where $h$, $r_1$ and $r_2$
%%are the cylinder heights, inner cylinder radius, and outer cylinder radius, respectively. 
%
The twist deformation is a fully three-dimensional deformation. 
(Specific Aim 3 address outstanding implementation issues like three-dimensional boundary integral solvers.) 
It is closely related to the assumption of lateral fluidity in membranes \cite{Hamm2000}. Note that tilt is locally non-zero whenever twist is non-zero.
That is because the surface gradient $\nabla {\bf n}$ equals the second fundamental form, and is therefore symmetric, whenever ${\bf n} = {\bf N}$ locally.

The proposed particle-based simulations of the HARP model in equations~\ref{SGD} also allow direct examination on the assumption of quadratic form of the elastic energy density.
As discussed earlier, recent disagreement in elastic moduli between continuum model and MD simulations originates from such assumption. 
The particle-based HARP model provides a unique opportunity to examine the validity of the quadratic elastic energy, and the PIs propose to examine how (if) any deviation from the quadratic elastic energy may depend on the microscopic details of the amphiphilic particles and their interactions.

%To measure the twist modulus, we place a single monolayer on a hydrophobic substrate 
%and take the initial configuration ${\bf n} = (0, f(x_1), \sqrt{1 - f^2(x_1)})$,
%yielding a surface gradient $\nabla {\bf n} = \begin{pmatrix}0 & 0\\ f'(x_1) & 0\end{pmatrix}$. 
%  Here $\Div {\bf n} = \mathrm{tr}(\nabla {\bf n}) = 0$, $\det(\nabla {\bf n}) = 0$ and
%  $\Curl {\bf n} = (\nabla u)_{12} - (\nabla u)_{21} = f'(x_1)$.
%  This gives the continuum elastic energy equals
%  \begin{equation}
%    \label{eq:twist_energy}
%    E = \int_{\Sigma} \tfrac{1}{2}\KT (f'(x_1))^2 + \tfrac{1}{2} \KTH f^2(x_1) \,\dif dx_1 \dif x_2,
%  \end{equation}
%  which we compare to the HAP simulation data.

\subsubsection{Asymptotic analysis on the limit $\rho\rightarrow 0$}

@@

Analogously, the HAP energy converges to the surface energy in the limit of vanishing screening length 
and the boundary layers of the minimizers (the solutions to the screen Laplace equation) converges to 
area measures supported on the particle boundaries. 
PI RR has worked directly on phase field approximations of membrane bending energy 
\cite{0951-7715-18-3-016} and asymptotcs in the Poisson-Boltzman equation \cite{1531-3492_2006_2_357,Lee2018}.
The proposed analysis is a natural outgrowth or prior work because the linearization of the Poisson-Boltzmann equation is the screened Laplace equations \eqref{SL}.
One of the basic questions is to establish the HAP convergence rate as a function of particle geometry 
\cite{LuMo89}, which is an essential benchmark for numerical validation. 
In the context of microstructure fabrication, the particle configurations maximally sequester hydrophobic surfaces.  
As an illustration, consider two fluid-bound plates $P_1$ and $P_2$ that geometrically flat and hydrophobic all sides.
We conjecture that in the zero-screening length limit, the plates have overlapping faces, and that 
\begin{equation}
\lim_{\rho \to 0} \min \Phi_{\text{hydro}} = \max_{\mathcal{X}} A(P_1 \Delta P_2)
\end{equation}
$A(P_1 \Delta P_2)$ is the area of the symmetric difference of the overlap.  
\cite{https://doi.org/10.1038/s41467-019-09787-6}
